---
title: "Reproduced analysis of FARS data"
author: "Ethan Walker"
date: "November 8, 2017"
output: pdf_document
fontsize: 10pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, include=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggthemes)

load("../data/clean_fars.RData")
source("../R/fars_functions.R")
```

#Results
  
##Percentages of drivers testing positive by drug type, sex, and year group
```{r}
table_one <- clean_fars %>%
  mutate(year_cat = cut(year, breaks = c(1999, 2002, 2006, 2010), 
                        labels = c("1999-2002", "2003-2006", "2007-2010"), 
                        include.lowest = TRUE, right = TRUE)) %>%
  filter(!is.na(sex)) %>%
  group_by(drug_type, sex, year_cat) %>%
  summarize(n_non_missing = sum(!is.na(positive_for_drug)), 
            positive_test = sum(positive_for_drug, na.rm = TRUE), 
            perc_positive = round(100 * positive_test / n_non_missing, 1)) %>%
  select(drug_type, sex, year_cat, perc_positive) %>%
  unite(sex_year_cat, sex, year_cat) %>%
  spread(sex_year_cat, perc_positive) %>%
  knitr::kable(col.names = c("Drug type", "F 1999-2002", "F 2003-2006", 
                             "F 2007-2010", "M 1999-2002", "M 2003-2006", 
                             "M 2007-2010"))
table_one
```

##Figure 1: Prevalence of nonalcohol drugs in fatally injured drivers by year and age group
```{r, fig.height=4, fig.width=7}
fig_one <- clean_fars %>% 
  filter(!is.na(agecat)) %>% 
  filter(!is.na(positive_for_drug)) %>% 
  filter(drug_type != "Alcohol") %>% 
  group_by(unique_id, year, agecat) %>% 
  summarize(positive_test = sum(positive_for_drug), 
            positive = any(positive_test)) %>% 
  ungroup() %>% 
  group_by(year, agecat) %>% 
  summarize(total_tests = length(positive), positive = sum(positive), 
            perc_positive = (100 * (positive / total_tests)))

fig_one_plot <- fig_one %>% 
  ggplot(aes(x = year, y = perc_positive, position = agecat)) +
  geom_line() +
  geom_point(aes(x = year, y = perc_positive, shape = agecat)) +
  scale_shape_manual("Age", values = c(16, 17, 15, 3)) +
  scale_y_continuous(limits = c(0, 35)) +
  theme_few() +
  labs(x = "Year", y = "Positive for Nonalcoholic Drugs, %")
fig_one_plot
```

##Figure 2: Prevalence of nonalcohol drugs in fatally injured drivers by year and drug type
```{r, fig.height=4, fig.width=7}
fig_two <- clean_fars %>% 
  filter(!is.na(positive_for_drug)) %>% 
  filter(drug_type != "Alcohol") %>% 
  group_by(year, drug_type) %>% 
  summarize(n_non_missing = sum(!is.na(positive_for_drug)),
    positive_test = sum(positive_for_drug, na.rm = TRUE),
    perc_positive = round(100 * positive_test / n_non_missing, 1))

fig_two_plot <- fig_two %>% 
  ggplot(aes(x = year, y = perc_positive, position = drug_type)) +
  geom_line() +
  geom_point(aes(x = year, y = perc_positive, shape = drug_type)) +
  scale_shape_manual("Drug Type", values = c(16, 17, 15, 3, 7)) +
  scale_y_continuous(limits = c(0, 13), breaks = c(0, 4, 8, 12)) +
  theme_few() +
  labs(x = "Year", y = "Positive for Drugs, %")
fig_two_plot

```

##Figure 3: Prevalence of cannabinoid drugs in fatally injured drivers by year and age group
```{r, fig.height=4, fig.width=7}
fig_three <- clean_fars %>% 
  filter(!is.na(agecat)) %>% 
  filter(drug_type == "Cannabinoid") %>% 
  group_by(year, agecat) %>% 
  summarize(n_non_missing = sum(!is.na(positive_for_drug)),
    positive_test = sum(positive_for_drug, na.rm = TRUE),
    perc_positive = round(100 * positive_test / n_non_missing, 1))

fig_three_plot <- fig_three %>% 
  ggplot(aes(x = year, y = perc_positive, position = agecat)) +
  geom_line() +
  geom_point(aes(x = year, y = perc_positive, shape = agecat)) +
  scale_shape_manual("Age", values = c(16, 17, 15, 3)) +
  scale_y_continuous(limits = c(0, 19), breaks = c(0, 5, 10, 15)) +
  theme_few() +
  labs(x = "Year", y = "Positive for Cannabinoid, %")
fig_three_plot

```

##Prevalence of drugs in fatally injured drivers for 1999 and 2010 by drug type
```{r}
prev_one <- clean_fars %>%
      filter(year %in% c("1999", "2010")) %>%
      group_by(drug_type, year) %>%
      summarize(positive = sum(positive_for_drug, na.rm = TRUE),
                trials = sum(!is.na(positive_for_drug)))

prev_two <- data.frame(prevalence = perc_cis(x = prev_one$positive, 
                                             n = prev_one$trials))

prev_one <- tibble::rowid_to_column(as.data.frame(prev_one, var = "rowid"))
prev_two <- tibble::rowid_to_column(as.data.frame(prev_two, var = "rowid"))

prev_final <- prev_one %>% 
  full_join(prev_two, by = "rowid") %>% 
  select(drug_type, year, prevalence) %>% 
  spread(key = year, value = prevalence) %>% 
  rename(`Drug type` = drug_type)
knitr::kable(prev_final)
```

##Statistics for testing for trend in prevalence of drugs over study years by drug type using Cochran-Armitage trend test
```{r}
drug_list <- c("Alcohol", "Nonalcohol", "Narcotic", "Depressant",
               "Stimulant", "Cannabinoid", "Other")
drug_trend_tests_ca <- lapply(drug_list, test_trend_ca)
drug_trend_tests_ca <- dplyr::bind_rows(drug_trend_tests_ca) %>%
  dplyr::mutate(drug = drug_list) %>%
  dplyr::select(drug, Z, p.value)
drug_trend_tests_ca %>% knitr::kable()
```

##Statistics for testing for trend in prevalence of drugs over study years by drug type using Wald test of logistic regression coefficient for “year”
```{r}
drug_list <- c("Alcohol", "Nonalcohol", "Narcotic", "Depressant",
               "Stimulant", "Cannabinoid", "Other")
drug_trend_tests_log_reg <- lapply(drug_list, test_trend_log_reg)
drug_trend_tests_log_reg <- dplyr::bind_rows(drug_trend_tests_log_reg) %>%
  dplyr::mutate(drug = drug_list) %>%
  dplyr::select(drug, Z, p.value)
drug_trend_tests_log_reg %>% knitr::kable()
```